TacticToe: Learning the Game of Tactics

### 2018-01-23 tactictoe_jv.tex
related works: wip

We bridge this gap by implementing our tactical prover TacticToe on top 
of the HOL4 ITP. 
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
% are effective from 
% tactic-level proofs and what kind of formulas are the results of effective 
% steps, 
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths. 
Experiments on the \holfour standard 
library show that TacticToe is able to re-prove 66.2\% of all 7164 theorems in 
less 
than 60 seconds. 
This is to be compared with the 42.9\% proved in the same amount of time by 
Eprover in auto-schedule mode with 128 lemmas predicted and 
translated by HolyHammer.
\todo{run experiments including proof scripts containing let
     run experiments without prove on 1/10th of the library.}

@@

\paragraph{Bound variables}
binding variables are not globalized

@@

\begin{definition}(Code of infix guards)
\begin{lstlisting}
fun a infixlX_open f = fn x ==> f (a,x)      fun g infixlX_close b = g b
fun a infixrX_open g = g a      fun f infixrX_close b = fn x ==> f (x,b) 
\end{lstlisting}
\end{definition}

@@

\begin{example} (Globalization)
\begin{lstlisting}
Tactic.STRIP_TAC 
  infixl0_open boolLib.THEN infixl0_close
Prim_rec.INDUCT_THEN (DB.fetch "list" "list_INDUCT") Tactic.ASSUME_TAC
  infixl0_open boolLib.THEN infixl0_close
Rewrite.ASM_REWRITE_TAC [DB.fetch "list" "MAP", DB.fetch "list" "APPEND"]
\end{lstlisting}
\end{example}

@@

We now restrict the definition of a proof script to being the third 
argument of a call to \texttt{store\_thm}. And we precisely define the 
modification applied to the theory file. A modified theory file is identical to 
the original theory file except every proof scripts \texttt{p} is
replaced by \texttt{Split "p'"} where \texttt{Split} is the splitting function 
to be defined and \texttt{"p'"} is the string representation of the globalized 
version of the proof script 
\texttt{p}. 
.
\begin{example}(Proof script transformation)
\begin{alltt}
original declaration: val id = store_thm (name, term, p)
modified declaration: val id = store_thm (name, term, Split "p'")
\end{alltt}
\end{example}

The intended effect of the splitting function is to detect and wrap \sml 
expressions of type tactics. Its aim is to is encapsulate each tactic 
unit $t$ within a recording function $R$. This function $R$ is designed to 
observe what goals $t$ receives at run time. Since there can be maybe nested 
level of tactics, (i.e 
tactics that contain tactics as subexpressions), we choose one that is
close to what a human understand to be a tactic unit. 
Most tacticals are infix, and we do not want to consider combination of 
tactics to be a tactic unit. For our wrapper \texttt{Split}, a tactic unit is a 
subterm of 
type tactics that do not contain a infix operator as its top symbol. Thanks to
globalized infixity, we just have to verify that the expression 
is of type tactic and that an infix guard is not a top symbol of the expression.
To determine that, we run the \polyml compiler on \texttt{"p'"} to give us a 
detailed code tree of the expression. Each tactic unit is then substituted 
by a wrapped version to produce the final proof script.

\begin{example} (Wrap)
\begin{alltt}\small
R "Tactic.STRIP_TAC"
  infixl0_open boolLib.THEN infixl0_close
R "Prim_rec.INDUCT_THEN (DB.fetch \bq{list}\bq{INDUCT}) Tactic.ASSUME_TAC"
  infixl0_open boolLib.THEN infixl0_close
R "Rewrite.ASM_REWRITE_TAC 
     [DB.fetch \bq{list} \bq{MAP}, DB.fetch \bq{list} \bq{APPEND}]"
\end{alltt}
\end{example}

@@

\begin{example}\label{ex:record} (Code of the recording function)
\begin{lstlisting}
fun R tactic_string goal = 
  (save (tactic_string, goal); (read_tactic tactic\_string) goal)
\end{lstlisting}
\end{example}

\subsection{Storing Training Data}
We describe here the format preferred to store and load the 
data recorded by the function \texttt{save}. 
During saving, we add fields to the input pair $(tactic,goal)$ to record 
precomputed information in order to make the prediction faster. The final 
feature vector is the tuple $(tactic,time,goal,goal\_list,features)$.
The time $time$ taken by $tactic$ is not currently used by our 
predictor. The list of goals $goal list$ is the output produced by $tactic$ on 
$goal$. And the features $features$ of the goal $goal$ are extracted and hashed 
as described in Section~\ref{sec:features}.

\begin{remark}
To simplify the description of our machine learning algorithms in future 
sections, we consider a feature vector to be a pair $(tactic,goal)$ and only 
explicitly rely on the 
other fields when necessary.
\end{remark}

@@

All top-level theorems in the standard library are considered except:
\begin{itemize}
\item the 619 calls to store\_thm that contains a let binding, these are 
typically 
harder for \tactictoe as it may not have seen the particular local declaration 
before and therefore would have to find an alternative proofs. We also do not 
learn from these proofs because they were explictly local.
\item the 1175 calls to save\_thm. These are typically mostly trivial and easy 
theorems.
The remaining 7088 top-level theorems are evaluated in the full scale 
evaluation and the level of difficulty of these problem is assessed by a 
comparison with \eprover.
\end{itemize}

@@

\begin{table}[h!]
\centering\ra{1.3}
\small
\begin{tabular}{llll}
\toprule
  & & Small(273) & Big(7088) \\
\midrule
  & \eprover (auto-schedule + 128 premises) & 86 & 2069 (29.19\%)\\ 
%  & \tactictoe (without special care for metis) & 154 & 3868 (\\
  & \tactictoe (best parameters) & 187 & 4186 (59.06\%) \\
%  & \tactictoe (+ self-learning) & 187 (187) &  4201 (59.27\%) \\
\bottomrule
\end{tabular}
\caption{Experiments with a timeout of 10 seconds \label{tab:_param}}
\end{table}

@@

\section{Related Work}\todo{TG reorganise + other motivation?}
There are several essential components of our work that are comparable to 
previous approaches: tactic-level proof recording, tactic 
selection through machine learning techniques and automatic tactic-based proof 
search. Our work is also related to previous approaches that use machine 
learning to select premises for the ATP systems and guide ATP proof search 
internally.

For \hollight, the Tactician tool~\cite{DBLP:conf/sefm/Adams15} 
can transform a packed tactical proof into a series of interactive tactic 
calls. Its principal application 
was so far refactoring the library and teaching common proof techniques to new 
ITP users. In our work, the splitting of a proof into a sequence of tactics is 
essential for the
tactic recording procedure, used to train our tactic prediction module.

The system 
\textsf{ML4PG}~\cite{DBLP:journals/corr/abs-1212-3618,DBLP:journals/mics/HerasK14}

groups related proofs thanks to its clustering 
algorithms. It allows \coq users to inspire themselves from similar proofs and 
notice 
duplicated proofs. Our predictions comes from a much more detailed description 
of the open goal.
However, we simply create a single label for each tactic call whereas each of 
its
arguments is treated independently in \textsf{ML4PG}. 
Our choice is motivated by the k-NN algorithm already used in
\holyhammer for the selection of theorems.
%, which can be easly adapted to the selection of tactics but would be improved 
%by a 
%version that can learn to predict tactics and their arguments independently.

\textsf{SEPIA}~\cite{DBLP:conf/cade/GransdenWR15} is a powerful system able to 
generate
proof scripts from previous \coq proof examples.
Its strength lies in its ability to produce likely sequences 
of tactics for solving domain specific goals. It operates by creating a model 
for common sequences of tactics for a specific library.
This means that in order to propose the following tactic, only the previously 
called tactics
are considered.
%intuition tells us that it should not be the main factor
Our algorithm, on the other hand, relies mainly on the characteristics of the 
current goal 
to decide
which tactics to apply next. In this way, our learning mechanism has to 
rediscover why each 
tactic was applied for the current subgoals. It may lack some useful bias for 
common sequences 
of tactics, but is more reactive to subtle changes. Indeed, it can be trained 
on a large library and only tactics relevant to the current subgoal will be 
selected. 
Concerning the proof search, \textsf{SEPIA}'s %simple 
breadth-first search is replaced by an A*-algorithm which allows for heuristic 
guidance in 
the exploration of the search tree.
Finally, \textsf{SEPIA} was evaluated on three chosen parts (totaling 2382 
theorems) of the 
\coq library demonstrating that it globally outperforms individual \coq 
tactics. In contrast, we demonstrate the competitiveness of our system against 
the successful general-purpose hammers on the \holfour standard library (7902 
theorems).

Internal learning-based selection of tactical steps inside an ITP is analogous 
to internal learning-based selection of clausal steps inside ATPs such as 
\textsc{MaLeCoP}~\cite{malecop} and \textsc{FEMaLeCoP}~\cite{femalecop}. These 
systems
use the naive Bayes classifier to  select clauses for the extension steps in
tableaux proof search based on many previous proofs. Satallax~\cite{Brown2012a} 
can guide its
search internally~\cite{mllax} using a command classifier, which can estimate 
the priority of the 11 kinds of
commands in the priority queue based on positive and negative examples.
There are several essential components of our work that are comparable to 
previous approaches: tactic-level proof recording, tactic 
selection through machine learning techniques and automatic tactic-based proof 
search. Our work is also related to previous approaches that use machine 
learning to select premises for the ATP systems and guide ATP proof search 
internally.

@@

The system 
\textsf{ML4PG}~\cite{DBLP:journals/corr/abs-1212-3618,DBLP:journals/mics/HerasK14}

groups related proofs thanks to its clustering 
algorithms. It allows \coq users to inspire themselves from similar proofs and 
notice 
duplicated proofs. Our predictions comes from a much more detailed description 
of the open goal.
However, we simply create a single label for each tactic call whereas each of 
its
arguments is treated independently in \textsf{ML4PG}. 
Our choice is motivated by the k-NN algorithm already used in
\holyhammer for the selection of theorems.
%, which can be easly adapted to the selection of tactics but would be improved 
%by a 
%version that can learn to predict tactics and their arguments independently.

\textsf{SEPIA}~\cite{DBLP:conf/cade/GransdenWR15} is a powerful system able to 
generate
proof scripts from previous \coq proof examples.
Its strength lies in its ability to produce likely sequences 
of tactics for solving domain specific goals. It operates by creating a model 
for common sequences of tactics for a specific library.
This means that in order to propose the following tactic, only the previously 
called tactics
are considered.
%intuition tells us that it should not be the main factor
Our algorithm, on the other hand, relies mainly on the characteristics of the 
current goal 
to decide
which tactics to apply next. In this way, our learning mechanism has to 
rediscover why each 
tactic was applied for the current subgoals. It may lack some useful bias for 
common sequences 
of tactics, but is more reactive to subtle changes. Indeed, it can be trained 
on a large library and only tactics relevant to the current subgoal will be 
selected. 
Concerning the proof search, \textsf{SEPIA}'s %simple 
breadth-first search is replaced by MCTS which allows for heuristic 
guidance in 
the exploration of the search tree.
Finally, \textsf{SEPIA} was evaluated on three chosen parts (totaling 2382 
theorems) of the 
\coq library demonstrating that it globally outperforms individual \coq 
tactics. In contrast, we demonstrate the competitiveness of our system against 
the successful general-purpose hammers on the \holfour standard library.
A recent improvement on this system~\cite{} includes the possibility to replace 
theorems in tactics by ones belonging to the same cluster, independently of the 
current goal. We have the opposite approach, we forget about the original 
theorems in the tactic creating an more general abstract tactic
and only use the goal to guide predictions. Finding a way to combine those two 
approaches could be the key to stronger theorem predictions.

A similar effort for stronger automation is performed in \isabelle. Nagashima 
developed a proof strategy language PSL~\cite{NagashimaK17psl} which allows the 
user to 
program with regular tactics and combine them with tools such as 
\sledgehammer~\cite{sledgehammer10} or Nitpick~\cite{Nitpick10}. 
Selection of tactics from recorded proofs has yet to be implemented and now it 
is up to come up with a good meta-tactic.

%In \holfour, 
%a tactic is a just a \sml function of a special type and \holyhammer can 
%easily 
%be modified to be a tactic. So this meta-level is not 
%necessary, all programming can occur at the tactic level and even \tactictoe  
%is a tactic.

%* Theory Exploration (Moa Johanson+, ITP 2017)  I think that is more related 
%to conjeturing and do not think it fits here.

>>>>

We bridge this gap by implementing our tactical prover TacticToe on top 
of the HOL4 ITP. 
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
% are effective from 
% tactic-level proofs and what kind of formulas are the results of effective 
% steps, 
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths. 
Experiments on the \holfour standard 
library show that TacticToe is able to re-prove 66.0\% of 
all theorems proven by a tactical script in the standard library.
This is to be compared with the 34.5\% proved in the same amount of time by 
Eprover in auto-schedule mode with 128 lemmas predicted and 
translated by HolyHammer.

@@

\paragraph{Bound variables}
Binding variables are not globalized

@@

\begin{definition}(Code of infix guards)
\begin{lstlisting}[language=SMLSmall,frame=tb]
fun a infixlX_open f = fn x ==> f (a,x)      fun g infixlX_close b = g b
fun a infixrX_open g = g a      fun f infixrX_close b = fn x ==> f (x,b) 
\end{lstlisting}
\end{definition}

@@

\begin{example} (Globalization)
\begin{lstlisting}[language=SMLSmall,frame=tb]
Tactic.STRIP_TAC 
  infixl0_open boolLib.THEN infixl0_close
Prim_rec.INDUCT_THEN (DB.fetch "list" "list_INDUCT") Tactic.ASSUME_TAC
  infixl0_open boolLib.THEN infixl0_close
Rewrite.ASM_REWRITE_TAC [DB.fetch "list" "MAP", DB.fetch "list" "APPEND"]
\end{lstlisting}
\end{example}

@@

We now restrict the definition of a proof script to being the third 
argument of a call to \texttt{store\_thm}. And we precisely define the 
modification applied to the theory file. A modified theory file is identical to 
the original theory file except every proof scripts \texttt{p} is
replaced by \texttt{Split "p'"} where \texttt{Split} is the splitting function 
to be defined and \texttt{"p'"} is the string representation of the globalized 
version of the proof script 
\texttt{p}. 
.
\begin{example}(Proof script transformation)
\begin{lstlisting}[language=SMLSmall,frame=tb]
original file: val id = store_thm (name, term, p)
modified file: val id = store_thm (name, term, Split "p'")
\end{lstlisting}
\end{example}

The intended effect of the splitting function is to detect and wrap \sml 
expressions of type tactics. Its aim is to is encapsulate each tactic 
unit $t$ within a recording function $R$. This function $R$ is designed to 
observe what goals $t$ receives at run time. Since there can be maybe nested 
level of tactics, (i.e 
tactics that contain tactics as subexpressions), we choose one that is
close to what a human understand to be a tactic unit. 
Most tacticals are infix, and we do not want to consider combination of 
tactics to be a tactic unit. For \texttt{Split}, a tactic unit is a 
subterm of 
type tactics that do not contain a infix operator as its top symbol. Thanks to
globalized infixity, we just have to verify that the expression 
is of type tactic and that an infix guard is not a top symbol of the expression.
To determine that, we run the \polyml compiler on \texttt{"p'"} to give us a 
detailed code tree of the expression. Each tactic unit is then substituted 
by a wrapped version to produce the final proof script.

\begin{example} (Wrap)
\begin{lstlisting}[language=SMLSmall,frame=tb]
R "Tactic.STRIP_TAC"
  infixl0_open boolLib.THEN infixl0_close
R "Prim_rec.INDUCT_THEN (DB.fetch \"list\" \"INDUCT\") Tactic.ASSUME_TAC"
  infixl0_open boolLib.THEN infixl0_close
R "Rewrite.ASM_REWRITE_TAC 
     [DB.fetch \"list\" \"MAP\", DB.fetch \"list\" \"APPEND\"]"
\end{lstlisting}
\end{example}

@@

\begin{example}\label{ex:record} (Code of the recording function)
\begin{lstlisting}[language=SMLSmall,frame=tb]
fun R tactic_string goal = 
  (save (tactic_string, goal); tactic_of_sml tactic_string goal)
\end{lstlisting}
\end{example}

\subsection{Storing Training Data}
We describe here the format preferred to store and load the 
data recorded by the function \texttt{save}. 
During saving, we add fields to the input pair $(tactic,goal)$ to record 
precomputed information in order to make the prediction faster. The final 
feature vector is the tuple $(tactic,time,goal,goal\_list,features)$.
The time $time$ taken by $tactic$ is not currently used by our 
predictor. The list of goals $goal list$ is the output of $tactic$ on 
$goal$. And the features $features$ of the goal $goal$ are extracted and hashed 
as described in Section~\ref{sec:features}.

\begin{remark}
To simplify the description of our machine learning algorithms in future 
sections, we consider a tactic feature vector to be a pair $(tactic,goal)$ and 
only 
explicitly rely on the 
other fields when necessary.
\end{remark}

@@

7000+ top-level theorems that were created by . 
In this evaluation, we do not include the 250 theorems created by

All top-level theorems created by a call to 
\texttt{store\_thm}, 
\texttt{Q.store\_thm} or \texttt{maybe\_thm} are considered with the exception 
of 440 calls that contains a let binding. They are typically harder to re-prove.
We also ignore the 1175 \texttt{save\_thm} calls. Most of them are easily 
provable theorems using instantiations or conjuncts of previous 
theorems.

%\begin{itemize}
%%\item the , these are 
%%typically 
%%harder for \tactictoe as it may not have seen the particular local 
%%declaration 
%%before and therefore would have to find an alternative proofs. We also do not 
%%learn from these proofs because they were explictly local.
%%\item the 1175 calls to save\_thm. These are typically mostly trivial and 
%%easy 
%%theorems.
%%The remaining 7088 top-level theorems are evaluated in the full scale 
%%evaluation and the level of difficulty of these problem is assessed by a 
%%comparison with \eprover.
%\end{itemize}

@@

\begin{table}[h!]
\centering\ra{1.3}
\small
\begin{tabular}{llll}
\toprule
  & & Small(273) & Big(7088) \\
\midrule
  & \eprover & 86 & 2069 (29.19\%)\\ 
%  & \tactictoe (without special care for metis) & 154 & 3868 (\\
  & \tactictoe & 187 & 4186 (59.06\%) \\
%  & \tactictoe (+ self-learning) & 187 (187) &  4201 (59.27\%) \\
\bottomrule
\end{tabular}
\caption{Experiments with a timeout of 10 seconds \label{tab:_param}}
\end{table}

@@

\section{Related Work}
There are several essential components of our work that are comparable to 
previous approaches: tactic-level proof recording, tactic 
selection through machine learning techniques and automatic tactic-based proof 
search. Our work is also related to previous approaches that use machine 
learning to select premises for the ATP systems and guide ATP proof search 
internally.

@@

The system 
\textsf{ML4PG}~\cite{DBLP:journals/corr/abs-1212-3618,DBLP:journals/mics/HerasK14}
groups related proofs thanks to its clustering 
algorithms. It allows \coq users to inspire themselves from similar proofs and 
notice 
duplicated proofs. Our predictions comes from a much more detailed description 
of the open goal.
However, we simply create a single label for each tactic call whereas each of 
its
arguments is treated independently in \textsf{ML4PG}. 
Our choice is motivated by the k-NN algorithm already used in
\holyhammer for the selection of theorems.
%, which can be easly adapted to the selection of tactics but would be improved 
%by a 
%version that can learn to predict tactics and their arguments independently.

\textsf{SEPIA}~\cite{DBLP:conf/cade/GransdenWR15} is a powerful system able to 
generate
proof scripts from previous \coq proof examples.
Its strength lies in its ability to produce likely sequences 
of tactics for solving domain specific goals. It operates by creating a model 
for common sequences of tactics for a specific library.
This means that in order to propose the following tactic, only the previously 
called tactics
are considered.
%intuition tells us that it should not be the main factor
Our algorithm, on the other hand, relies mainly on the characteristics of the 
current goal 
to decide
which tactics to apply next. In this way, our learning mechanism has to 
rediscover why each 
tactic was applied for the current subgoals. It may lack some useful bias for 
common sequences 
of tactics, but is more reactive to subtle changes. Indeed, it can be trained 
on a large library and only tactics relevant to the current subgoal will be 
selected. 
Concerning the proof search, \textsf{SEPIA}'s %simple 
breadth-first search is replaced by MCTS which allows for supervised learning
guidance in the exploration of the search tree.
Finally, \textsf{SEPIA} was evaluated on three chosen parts of the 
\coq library demonstrating that it globally outperforms individual \coq 
tactics. In contrast, we demonstrate the competitiveness of our system against 
the ATP \eprover on the \holfour standard library.
A recent improvement on this system~\cite{}\todo{Cezary: could not remember the 
reference that you gave me here i think it was CICM 2017} includes the 
possibility to 
substitute theorems appearing in tactics by ones belonging to the same cluster. 
In this work, we are also able to predict theorems appearing as arguments of 
tactics but we rely only on information from the current goal. Combining the 
two approaches by predicting from the goal and from the original tactic 
arguments is more difficult but is likely to produce more suitable lemmas.

A similar effort for stronger automation is performed in \isabelle. Nagashima 
developed a proof strategy language PSL~\cite{NagashimaK17psl} which allows the 
user to 
program with regular tactics and combine them with tools such as 
\sledgehammer~\cite{sledgehammer10} or Nitpick~\cite{Nitpick10}. 
Selection of tactics from data collected in recorded proofs is in development 
but for now it is up to the user to come up with a good meta-tactic.
